Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-15 paper.tex

@@

@@

\section*{Introduction}

@@

\section*{Methods}

@@

\subsection*{Recombination}

@@

\subsection*{Ancestral recombination graphs}

@@

\subsection*{Gene conversion}

@@

\subsection*{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are provided, which allow for
varying population parameters and migration rates over time, as well as
population splits, admixtures and pulse migrations. The location of lineages
over time can be tracked in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.

@@

\subsection*{Instantaneous bottlenecks}

@@

\subsection*{Multiple merger coalescents}

@@

\subsection*{Selection}
Selection was added to the coalescent ... Lots of simulators exist:
SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs},
msms~\citep{ewing2010msms},
% What about Cosi, did it simulate selection?
Cosi2~\citep{shlyakhter2014cosi2},
and discoal~\citep{kern2016discoal}.

@@

\subsection*{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

@@

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection*{Simulation interface}

@@

\subsection*{Data interchange and interoperability}
The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important---if often overlooked---issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

@@

\subsection*{Development model}
The community development process for \msprime\ is enabled by following an
open source development practise with a strong emphasis on code quality
and correctness. To ensure a consistent style, we require that all code
follows the PEP8 guidelines. Unit tests for all additions are required, and
are automatically run on each pull request on a variety of continuous
integration testing platforms. For new simulation features, statistical tests
must also be added, comparing the distribution of results to either analytical
results or existing simulators.

@@

\section*{Results and Discussion}

\subsection*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to the seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and
intricate algorithms~\citep{kelleher2016efficient}.

@@

\section*{Acknowledgments}
JK is supported by the Robertson Foundation.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
[Others fill in their ACKs please]

@@

>>>>

\usepackage[]{hyperref}
\hypersetup{
    pdftitle={Msprime 1.0: an efficient and extensible coalescent simulation framework},
    pdfauthor={Msprime developers},
    pdfsubject={Coalescent simulation},
    pdfkeywords={coalescent, mutation, simulation, genetic ancestry, genealogy, evolution},
    bookmarksnumbered=true,     
    bookmarksopen=true,
    bookmarksopenlevel=3,
    colorlinks=true,
    pdfstartview=Fit,
    pdfpagemode=UseOutlines,
    pdfpagelayout=TwoPageRight
}

@@

\setcounter{secnumdepth}{0} %Do not print out section numbers

@@

\section{Introduction}

@@

\section{Methods}

@@

\subsection{Recombination}

@@

\subsection{Ancestral recombination graphs}

@@

\subsection{Gene conversion}

@@

\subsection{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are provided, which allow for
varying population parameters and migration rates over time, as well as
population splits, admixtures and pulse migrations. The location of lineages
over time can be tracked in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.

@@

\subsection{Instantaneous bottlenecks}

@@

\subsection{Multiple merger coalescents}

@@

\subsection{Selection}
Selection was added to the coalescent ... Lots of simulators exist:
SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs},
msms~\citep{ewing2010msms},
% What about Cosi, did it simulate selection?
Cosi2~\citep{shlyakhter2014cosi2},
and discoal~\citep{kern2016discoal}.

@@

\subsection{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

@@

\subsection{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection{Simulation interface}

@@

\subsection{Data interchange and interoperability}
The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important---if often overlooked---issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

@@

\subsection{Development model}
The community development process for \msprime\ is enabled by following an
open source development practise with a strong emphasis on code quality
and correctness. To ensure a consistent style, we require that all code
follows the PEP8 guidelines. Unit tests for all additions are required, and
are automatically run on each pull request on a variety of continuous
integration testing platforms. For new simulation features, statistical tests
must also be added, comparing the distribution of results to either analytical
results or existing simulators.

@@

\section{Results and Discussion}

\subsection{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to the seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and
intricate algorithms~\citep{kelleher2016efficient}.

@@

\section{Acknowledgments}
JK is supported by the Robertson Foundation.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
[Others fill in their ACKs please]

@@

\setcounter{secnumdepth}{2} % Print out appendix section numbers
